---
title: "Introduction to data vizualisation - Part 2"
output_dir: "C:/Users/benjamcr/Rproj/GEOG3006/docs"
output: 
      distill::distill_article:
        self_contained: false
        toc: TRUE
        toc_float: TRUE
        number_sections: yes
---

## Exercise correction

First of all, we need to open the dataset. 

```{r, echo = TRUE}
nobel <- read.csv('C:/Users/benjamcr/Rproj/GEOG3006/data/nobel.csv')
head(nobel)
```


<span style="color: blue;">**Question 1**</span>: How many observations and how many variables are in the dataset?

We can use the functions `nrow` and `ncol` to get the number of observations and the number of variables in a dataset. As you may remember from the tutorials, the **rows** of a dataset are the **observations** (Tor, Gus or Lena in the `kids_frame` dataset for instance). The **columns** are the variables (names, shirt_color or height in the `kids_frame` dataset)

```{r, echo = TRUE}
# Number of observations
nrow(nobel)

# Number of variables
ncol(nobel)
```


<span style="color: blue;">**Question 2**</span>: How many woman won a nobel price? How many men?

Although there is multiple ways of answering this question in R, a fast and efficient way of doing so is to use the combination `group_by` and `summarise()`, functions included in the `dplyr` package.

```{r, echo = TRUE}
library(tidyverse)

nobel %>% 
  dplyr::group_by(gender) %>% 
  dplyr::summarise(n = n())
  
```


> In the line above I load the library `tidyverse`. Later in the document we will see what is this library and why it is so useful.

From the output we can see that only 52 women got a nobel price while 856 men got a nobel price. 

We can calculate the percentage of women who got a nobel price:

```{r}
52 / (52 + 856)
```

Only 5%!

<span style="color: blue;">**Question 3**</span>: Create a new data frame called nobel_living that filters for

* laureates for whom country is available (you can use the function `drop_na()`, look on google what it does!)
* females laureates who are still alive (their died_date is NA)

Here, we need to create a new object that we will call `nobel_living`. To select the data we are supposed to place in this dataset we will use the function `filter`, again from the `dplyr` package.

```{r, echo = TRUE}
nobel_living <- nobel %>% 
  drop_na(country) %>% 
  filter(gender == "female") %>% 
  filter(is.na(died_date))
```

The code above has three steps. Note that in this case there is no special order for the steps.

* First, I filter only for the laureates whom country is available with the function `drop_na`
* Then I filter only for laureates whom gender is `female`
* Finally, I filter only for the laureates whom died_date is `NA`. Here I combine 2 functions: `filter` and `is.na()`.

The function `is.na()` will test whether a particular cell value is `NA` or not. Here, if NA is true (the person is still living) then the function will filter for it. Roughly, you can translate this line of code by **Filter the laureates whom died_date is NA**.

> Note that you can filter only the dead persons by adding a ! before the `is.na()` function.

```{r, echo = TRUE, eval =FALSE}
nobel_living <- nobel %>% 
  drop_na(country) %>% 
  filter(gender == "female") %>% 
  filter(is.na(died_date))
```

We can translate this as **Filter the laureates whom died_date is no NA**.

<span style="color: blue;">**Question 4**</span>: With this new dataset, summarize the number of females laureate who are still alive by country and make a histogram of the number of female laureates per country. Your histogram should include a title and title for the axis.

First, we will create a dataset which summarise the number of female laureates by country:

```{r, echo = TRUE}
nobel_female_country <- nobel_living %>% 
  group_by(country) %>% 
  summarise(n = n())
```

This dataset should have 7 observations and only two variables which are `n` the number of living female laureate per country and `country`.

Then we can build our visualisation with the package `ggplot2`. Here I do not build a histogram but rather a **lollipop chart**.

```{r, echo = TRUE}
nobel_female_country %>% 
  ggplot(., aes(x = country, y = n)) +
  geom_point(size = 3, color = "#a25079") +
  geom_segment(aes(x = country, xend = country, y=0, yend = n), color = "grey") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Number of nobel prices per country",
       xlab = "Number of nobel prices",
       ylab = "Country")
```


<span style="color: blue;">**Question 5**</span>: Which country has the most female laureates?

From the lollipop chart we draw we can clearly see that the US has the most female living laureate. Note that Norway has a woman who won a Nobel Price and who is still alive, in fact **May-Britt Moser** works at NTNU in St-Olav!


## A word on the `tidyverse` library

Previously, I have used the [Tidyverse](https://www.tidyverse.org/#:~:text=Tidyverse&text=The%20tidyverse%20is%20an%20opinionated,%2C%20grammar%2C%20and%20data%20structures.&text=See%20how%20the%20tidyverse%20makes,%E2%80%9CR%20for%20Data%20Science%E2%80%9D.) package which is a **collection** of R packages that share an underlying design philosophy, grammar and data structures.

We have already been using some of the packages included in the `tidyverse` including `dplyr`, `magritrr` (package which has the pipe) and `ggplot2`. There is a lot more to explore and a lot of very useful functions in it. Of course I do not know everything from the `tidyverse` and later in this course you may find more efficient ways to **wrangle** the data than I do, let me know if that's the case!

I highly recommend to install the `tidyverse` package and to load it at the beginning of your script. If you do that you won't have to load `ggplot`, `dplyr` ... every time.

```{r, echo = TRUE, eval = FALSE}
install.packages('tidyverse')
library(tidyverse)
```

If you want to improve your data science skills you can find some tips [here](https://oliviergimenez.github.io/tidyverse-tips/).

## Other functions that will make your life easier in R

You already have been introduced to some of the main functions from the `dplyr` package:

* `filter`
* `group_by`
* `summarise`

There is two other functions which will help you in your data analysis workflow, namely `select()` and `mutate()`

### A few words on `select()`

It’s not uncommon to get datasets with hundreds or even thousands of variables. In this case, the first challenge is often narrowing in on the variables you’re actually interested in. `select()` allows you to rapidly zoom in on a useful subset using operations based on the names of the variables.

`select()` is not very useful on the `kids_frame` dataset but you can still get the idea:

```{r, echo = TRUE}
# We reconstruct the kids_frame dataset
kids_frame <- data.frame(
  names = c("Tor", "Gus", "Bob", "Di", "Lena", "Tony", "Ingrid", "Maria", "Ed", "Raghnild"),
  height = c(110, 130, 115, 140, 125, 135, 120, 130, 130, 115),
  shirt_color = c("green", "green", "green", "blue", "blue", "green", "blue", "green", "green", "blue"),
  shoe_color = c("blue", "red", "grey", "blue", "pink", "red", "grey", "pink", "pink", "blue"),
  sex = c("m", "m", "m", "f", "f", "m", "f", "f", "m", "f"),
  age = c(8,11,8,12,11,11,9,12,12,8))
```

Imagine you would like to create a new dataset with only the columns names and height, with the `select()` function you would write:

```{r, echo = TRUE}
kids_selected <- kids_frame %>% 
  dplyr::select(names, height)

head(kids_selected)
```

Now you reduced the initial dataset to only two columns.

### A few words on `mutate()`

Besides selecting sets of existing columns, it’s often useful to add new columns that are functions of existing columns. That’s the job of `mutate()`.

`mutate()` always adds new columns at the end of your dataset so we’ll start by creating a narrower dataset so we can see the new variables. Remember that when you’re in RStudio, the easiest way to see all the columns is `View()`

Here I will create a new column about the kids' favorite food:

```{r}
kids_frame_food <- kids_frame %>% 
  mutate(fav_food = c("strawberry", "candy", "pasta", "chocolate", "candy", "beef", "soup", "brocoli", "cake", "ice cream"))

head(kids_frame_food)
```

Now you can see that the new column has been created!

### Other functions worth mentioning

Now you know the main functions of the tidyverse package that will allow you to wrangle efficiently. The list is not exhaustive and as I said I do not know all the functions! However here a small list of the `tidyverse` functions worth mentionning. You can look them up in R by typing `?NAME_FUNCTION` or on google.

* `transmute`
* `rename`
* `starts_with`
* `ends_with`
* `slice`
* ...

### For next lab

During next lab we will finish the bloc on Data visualisation and data wrangling. We will mainly repeat what we have been doing, using the same functions. 

First, download the dataset here: 

```{r, echo = FALSE}
xfun::embed_file('C:/Users/benjamcr/Rproj/GEOG3006/data/ncbikecrash.rda')
```

As you may see, this is an **.rda** file. A **.rda** file is basically a compressed R file. I saved this file in the folder "data". TO open it I should write:

```{r, echo = TRUE, eval = FALSE}
load("data/ncbikecrash.rda")
```

You will use what you learned form the previous labs and answer these questions:

<span style="color: blue;">**Question 1**</span>: Run View(ncbikecrash) in your Console to view the data in the data viewer. What does each row in the dataset represent?

<span style="color: blue;">**Question 2**</span>: How many bike crashes were recorded in NC between 2007 and 2014? How many variables are recorded on these crashes?

<span style="color: blue;">**Question 3**</span>: How many bike crashes occurred in residential development areas where the driver was between 0 and 19 years old?

<span style="color: blue;">**Question 4**</span>: Create a frequency table of the estimated speed of the car (driver_est_speed) involved in the crash. What is the most common estimated speed range in the dataset?

Good luck! 