---
title: "Modelling with a single predictor - 2"
output_dir: "C:/Users/benjamcr/Rproj/GEOG3006/docs"
output: 
      distill::distill_article:
        self_contained: false
        toc: TRUE
        toc_float: TRUE
        number_sections: yes
---

# Correction exercise

First of all, we load the libraries and open our dataset:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
library(tidyverse)
library(openintro)
library(broom)
```

```{r, echo = TRUE}
data <- openintro::evals
```

### Part 1: Exploratory Data Analysis

<span style="color: #9B870C;">**Question 1:**</span> Visualize the distribution of score. Is the distribution skewed (i.e. does your data look **normally distributed** or are they concentrated towards certain values)? What does that tell you about how students rate courses? Is this what you expected to see? Why, or why not? Include any summary statistics and visualizations you use in your response.

First, we visualize the distribution of `score` by looking at its distribution. For that we can make a histogram. Here we can basically play around with `score` and visualise this variable against the for instance `age`, `gender` or `pic_outfit`. 

```{r, echo = TRUE}
ggplot(data, aes(x=score)) + geom_histogram() + theme_bw()
```

Even though `score` looks a bit skewed towards higher values, the shape of the histogram looks more or less **normally distributed**. This means that mainly the students grade the course generously. However, we need to look at the reasons explaining the low grades or the very high grades to get a more complete picture of the grading!

> Data analysis is like doing some detective work. Once we have detected a pattern we need to investigate the reasons of this pattern.

To investigate the pattern we can have a look at how is `score` distributed relatively to the other variables by either making a `boxplot` (if the other variable is **categorical** or a `scaterplot` (if the other variable is **numeric**). To provide some examples here I plot the distribution of score against `age`, `pic_outfit` or `gender` 

```{r, echo = TRUE}
a <- ggplot(data, aes(y=score, x= pic_outfit)) + geom_boxplot() + theme_bw ()
b <- ggplot(data, aes(y=score, x= age)) + geom_point() + theme_bw ()
c <- ggplot(data, aes(y=score, x= gender)) + geom_boxplot() + theme_bw ()
cowplot::plot_grid(a,b,c, nrow = 1)
```

To plot multiple `ggplot` objects we can use the function `plot_grid` from the library `cowplot`. Write `?plot_grid` to get more information on the function!

On these plots we can see that teachers wearing formal outfits are graded higher, men teachers are graded higher than women teachers ... It seems difficult to infer any trend for age - we would need to enhance the plot.

<span style="color: #9B870C;">**Question 2:**</span> Visualize and describe the relationship between **score** and **bty_avg**.

We make our plot of `score` versus `bty_avg` (diminutive of **beauty average**):

```{r, echo = TRUE}
ggplot(data, aes(y=score, x= bty_avg)) + geom_point()
```

It looks like `score` gets higher as `bty_avg` gets higher. Nevertheless the pattern does not look so clear when plotted as it is at the moment. We can do better `r emo::ji("east")`

<span style="color: #9B870C;">**Question 3:**</span>  Replot the scatterplot (the point cloud) from the previous question, but this time use `geom_jitter()`? What does “jitter” mean? What was misleading about the initial scatterplot?

To make the pattern appear clearer we can use `geom_jitter`, function which adds a small amount of variation to the location of each points. Thus, the points will not overlap anymore and we can get a clearer picture of the distribution of the points.

```{r, echo = TRUE}
ggplot(data, aes(y=score, x= bty_avg)) + geom_point() + geom_jitter()
```

Our intuition now gets stronger: the "prettier" the teacher, the higher the students will grade her / him.

### Part 2 - Linear regression with a numerical predictor

<span style="color: #9B870C;">**Question 4:**</span> Let’s see if the apparent trend in the plot is something more than natural variation. Fit a linear model called `m_bty` to predict average professor evaluation score by average beauty rating (bty_avg). Based on the regression output, write the linear model and interpret your model

We wish to "predict" or explain the variable `score` by the variable `bty_avg`. Thus, `score` is our **response variable** while `bty_avg` is a **predictor** or an **explanatory variable** (both terminology can be used, you will also encounter the word **covariate**).

Based on this, the equation of the model look like this:

$score = \beta_0 + \beta_1 * beauty average + \epsilon$

$\epsilon$ being some noise normally distributed - recall the lecture on the residual values!

The model we are running aims to evaluate the effect of `bty_avg` and therefore tries to estimate the $\beta_0$ and the $\beta_1$ of out equation. We call them the **parameters**. We can run the model in R using the function `lm()`:

```{r, echo = TRUE}
# Fit the linear model
m_bty <- lm(score ~ bty_avg, data = data)

# Get a look at the model
summary(m_bty)
```

If, like me, you find the `summary` of the model overwhelming you can use the function `tidy` from the `broom` package. This gives us a more structured output at the **parameters values**.

```{r, echo = TRUE}
broom::tidy(m_bty)
```

In the model output the `(Intercept)` represents the $\beta_0$ of our equation while the `bty_avg` represents the $\beta_1$ of the model.

Given the **estimated values of our parameters** (be careful about the vocabulary here!) we can re-write the equation previously defined:

$score = 3.88 + 0.0666 * beauty average + \epsilon$

This means that if the teacher has a `bty_avg` of 0, he will still get, on average, a grade of 3.88. Moreover, for 1 more unit of bty_avg, the grade of the teacher will increase by 0.0666. Our intuition is thus confirmed: **the prettier the teacher, the higher will be her / his grade on average**.


<span style="color: #9B870C;">**Question 5:**</span> Replot your visualization from **Question 3**, and add the regression line to this plot in orange color. Turn off the shading for the uncertainty of the line.

We have 2 ways of plotting the regression line. The easiest is to use the `geom_smooth` function from `ggplot2`. However, we need to tweak a bit its arguments. First, we want the line resulting from a **linear model** so we need to specify `model = 'lm'` in the function. By default, `geom_smooth` use another kind of model (a generalised additive model which is a fancier version of the linear model). Then we need to turn off the uncertainty of the line. We need the specify the argument `se = FALSE` (`se` stands for **standard error**).

```{r, echo = TRUE}
ggplot(data, aes(y=score, x= bty_avg)) + 
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE)

```

Another way of making the line is by "predicting" what the `score` would be based on the **parameter values** estimated by our linear model. Basically, the `predict` function will replace the $beautyaverage$ of out equation by its value and will compute the `score`.

For instance, based on the linear model I can predict what the `score` would be for a `bty_avg` value of 2:

$score = 3.88 + 0.0666 * beauty average$
$score = 3.88 + 0.0666 * 2$
$score = 4.0132$

The `predict` function does this automatically. In the code below, we create a new column `score_pred` with the `predict` function, then we **plot the data** in a first part and we **plot the regression line obtained through the predictions**.

```{r}
data$score_pred <- predict(m_bty, data) 

ggplot(data, aes(x= bty_avg, y = score)) + 
  geom_point() + 
  geom_jitter() +
  geom_line(aes(y = score_pred), color = 'red')
```

As you can see, both method gives the same output.

<span style="color: #9B870C;">**Question 6:**</span> Interpret the slope of the linear model in context of the data.

For a 1 unit increase in `bty_avg` the score will increase by 0.066 on average

<span style="color: #9B870C;">**Question 7:**</span> Interpret the intercept of the linear model in context of the data. Comment on whether or not the intercept makes sense in this context.

For a `bty_avg` of 0, the score will be of 3.88

$score = 3.88 + 0.0666 * beauty average$
$score = 3.88 + 0.0666 * 0$
$score = 3.88$


### Part 3: Linear regression with a categorical predictor

<span style="color: #9B870C;">**Question 8:**</span> Fit a new linear model called `m_gen` to predict average professor evaluation `score` based on `gender` of the professor. Based on the regression output, write the linear model and interpret the slope and intercept in context of the data.

First of all, we need to code `gender` as a **dummy variable**. This means that we will recode the categoris `male` and `female` into numbers. The `lm` function does this automatically. In my opinion and from my experience it is preferable to recode the variable yourself so the process is more transparent.

In the code below we recode `female` as 1 and `male` as 0. I use the function `ifelse` which takes 3 arguments:

* The condition: does X = Z?
* The outcome if the condition is true
* The outcome if the condition is false

We translate the code below as "if gender = famale then the value is 1, otherwise the value is 0".

```{r, echo = TRUE}
data$gender_recode <- ifelse(data$gender == 'female', 1, 0)
```

Now we can fit the model:

```{r, echo = TRUE}
m_gen <- lm(score ~ gender_recode, data = data)

# Generate a tidy output with the tidy function from the broom package:
broom::tidy(m_gen)
```

The model that we just fitted has the form:

$score = \beta_0 + \beta_1 * gender + \epsilon$

We can interpret the $\beta_0$ as the average `score` for male professor and $\beta_1$ the "additional value to score" if the professor is a female.


<span style="color: #9B870C;">**Question 9:**</span> What is the equation of the line corresponding to male professors? What is it for female professors?

We can formalise what we previously said by giving the equation for both male and females:

$score_{female} = \beta_0 + \beta_1 * 1$
$score_{female} = 4.2 - 0.14 * 1$
$score_{female} = 4.06$

$score_{male} = \beta_0 + \beta_1 * 0$
$score_{male} = 4.2 - 0.14 * 0$
$score_{male} = 4.2$

On average, the students will grade a male teacher 4.2 while they will grade a female professor 4.06.

# Exercise continuation

For this lab and the next one you will continue to analyze the `eval` dataset. Here, you will still work with **categorical predictors** and learn how to interpret it. It is very important to understand how to interpret models with categorical data, especially in the social sciences where you will be confronted to plenty of them. Here are the questions for the next lab:

<span style="color: #9B870C;">**Question 10:**</span> Fit a new linear model called `m_rank` to predict average professor evaluation score based on `rank` of the professor. Based on the regression output, write the linear model and interpret the slopes and intercept in context of the data.

> Hint: this time there is 3 categories ...

<span style="color: #9B870C;">**Question 11:**</span> Create a new variable called `rank_relevel` where `"tenure track"` is the baseline level.

> Hint: the baseline level would be the intercept ...


<span style="color: #9B870C;">**Question 12:**</span> Fit a new linear model called `m_rank_relevel` to predict average professor evaluation `score` based on `rank_relevel` of the professor. This is the new (releveled) variable you created in Exercise 13. Based on the regression output, write the linear model and interpret the slopes and intercept in context of the data. Also determine and interpret the **R2** of the model.


<span style="color: #9B870C;">**Question 13:**</span> Create another new variable called `tenure_eligible` that labels `"teaching"` faculty as "no" and labels `"tenure track"` and `"tenured"` faculty as `"yes"`.


<span style="color: #9B870C;">**Question 14:**</span> Fit a new linear model called `m_tenure_eligible` to predict average professor evaluation `score` based on `tenure_eligibleness` of the professor. This is the new (regrouped) variable you created in Exercise 15. Based on the regression output, write the linear model and interpret the slopes and intercept in context of the data. Also determine and interpret the **R2** of the model.